Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-05-04 RNARedPrint-Bioinfo2018.tex
Add PK targeting figure; replace deviation by linear fit figure; move pk-free figure to supplemental material

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rejection approach is justified by the following considerations:
i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist}).
%Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances.
For base pair-based energy models and special base pair dependency graph
(paths, cycles\ldots) this property rigorously follows from analytic
combinatorics, see \citet{Bender1983} and
\citet{Drmota1997}. In such cases, the expected number of
rejections before reaching the targeted energies remains constant when
$\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
$\varepsilon=0$. The \GCb\% of designs can also be controlled,
jointly and in a similar way, as done in
\Software{IncaRNAtion}~\citep{Reinharz2013}.

@@

Fig.~\ref{fig:energydist} illustrates the described
strategy. Fig.~\ref{fig:energydist}A
illustrates our capability to target Turner energies based on the
stacking energy model. It shows the deviation between
$\EnergyTurner$ and $\EnergyStacking$, observed for instances of the
\texttt{Modena} benchmark. For each structure, $n=1\,000$ sequences
were sampled, while targeting a ${\sf GC}\%$ of $50\%$, and reported
the difference between the expected and sampled energies for both the
Turner model and its approximation using a stacking model.  For an
example instance of the \texttt{Modena} benchmark \texttt{3str},
Fig.~\ref{fig:energydist}B shows the Turner energy distributions of
the single structures as they result from sampling with different weight
parameters. The figure illustrates how our multidimensionnal Boltzmann
sampling strategy can accurately and, to a large extent, independently
shift the Turner energies of sampled sequences to prescribed targets.

%To reduce the cost of the precomputation, we used a stacking energy model 
%as a proxy for the Turner model. Indeed, we observed a strong linear 
%dependency between the Turner energy and the 
%emphasize that only the simple stacking energies are directly targeted
%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
%linear adjustment, we finally achieve well-controlled energies in the
%realistic Turner energy model.
\begin{figure*}[t]
  \begin{center}
    {\sf \bfseries A}\includegraphicstop[width=0.32\textwidth]{Figs/offset2}\hspace{.5cm}
    {\sf \bfseries B}\includegraphicstop[width=0.57\textwidth]{Figs/energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies using multi-dimensional Boltzmann
    sampling.  (A) Accuracy of targeting over all target structures of
    the \Software{Modena} benchmark instances \texttt{RNAtabupath}
    (2str), \texttt{3str} and \texttt{4str}. 
    For the stacking and the Turner energy model, the relative offsets between the targeted and 
    achieved energies were determined. We show the deviations of these offsets between both models.
    (B) Turner energy distributions for
    three target structures S1, S2 and S3. We show the results of
    targeting the respective free energies $(-40,-40,-20)$ and
    $(-20,-20,-20)$ for the target structures (annotated in red and
    blue, resepectively)---this demonstrates the
    effectivity of our adaptive multi-dimensional Boltzmann sampling
    procedure; moreover, for comparison, the distributions for uniform
    (U) and Boltzmann (B) samples---respectively associated with
    homogenous weights $1$ and $e^\beta$.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
We empirically study the capacity of \ourprog{} to improve the generation of seed sequences for subsequent local optimizations, showing an important application in biologically relevant scenarii.

>>>>

A further \Def{rejection step} is applied to the generated structures to retain only sequences whose energy for each structure $R_\ell$ belongs to $[\TargetE_\ell\cdot(1-\varepsilon),\TargetE_\ell\cdot(1+\varepsilon)]$, for $\varepsilon\ge 0$ some predefined \Def{tolerance}. The rejection approach is justified by the following considerations:
i) \emph{Enacting an exact control over the energies would  be technically hard and costly.} Indeed, controlling the energies through dynamic programming would require explicit convolution products, generalizing~\citet{Cupal1996}, inducing additional $\Theta(n^{2k})$ time and $\Theta(n^k)$ space overheads;
ii) \emph{Induced distributions are typically concentrated.} Intuitively, unless sequences are fully constrained individual energy terms are independent enough so that their sum is concentrated around its mean -- the targeted energy (cf Fig.~\ref{fig:energydist-pk}).
%Indeed, as shown in Fig.~\ref{fig:shifting_mean}, empirical joint distributions in the energies have good fits towards Normal multidimensional distributions with linear (co-)variances.
For base pair-based energy models and special base pair dependency graph
(paths, cycles\ldots) this property rigorously follows from analytic
combinatorics, see \citet{Bender1983} and
\citet{Drmota1997}. In such cases, the expected number of
rejections before reaching the targeted energies remains constant when
$\varepsilon\ge 1/\sqrt{n}$, and $\Theta(n^{k/2})$ when
$\varepsilon=0$. The \GCb\% of designs can also be controlled,
jointly and in a similar way, as done in
\Software{IncaRNAtion}~\citep{Reinharz2013}.

@@

Fig.~\ref{fig:energydist-pk} illustrates the described strategy. For
the two target structures of Fig.~\ref{fig:energydist-pk}B,
Fig.~\ref{fig:energydist-pk}A shows the good fit between Turner
energies based and energies in the stacking energy model. For the
shown fits we sampled $n=1\,000$ sequences, targeting a ${\sf GC}\%$
of $50\%$.
%
For an example instance of the \texttt{Modena} benchmark with two
pseudoknotted target structures, Fig.~\ref{fig:energydist-pk}B shows
the Turner energy distributions of the single structures as they
result from sampling with different weight parameters. The figure
illustrates how our multidimensionnal Boltzmann sampling strategy can,
to a large extent, independently shift the Turner energies of sampled
sequences towards prescribed targets. See Supp. Mat.,
Fig.~\ref{fig:energydist} for a further example with three
pseudoknot-free target structures.

%To reduce the cost of the precomputation, we used a stacking energy model 
%as a proxy for the Turner model. Indeed, we observed a strong linear 
%dependency between the Turner energy and the 
%emphasize that only the simple stacking energies are directly targeted
%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
%linear adjustment, we finally achieve well-controlled energies in the
%realistic Turner energy model.
\begin{figure*}[t]
  \begin{center}
    {\sf \bfseries A}\ \includegraphicstop[width=0.38\textwidth]{Figs/regression-pk}\hfill
    {\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/energy_distribution-pk}
  \end{center}
  \caption{%
    Targeting specific energies for pseudoknotted structures using multi-dimensional Boltzmann
    sampling.  (A) Linear fits between the energies in the stacking model to the Turner model for
    sampled sequences; the good match enables more efficient targeting of Turner energies based on
    targeting stacking model energies.
    We show the fits for both target structures S1 and S2 of subfigure B.
    (B) Turner energy distributions for
    two target structures S1 and S2. We show the results of
    targeting the respective free energies $(-20,-10)$, $(-10,-20)$, and $(-15,-15)$
    for the target structures (annotated in red,
    yellow, and blue resepectively)---this demonstrates the
    effectivity of our adaptive multi-dimensional Boltzmann sampling
    procedure; moreover, for comparison, the distributions for uniform
    (U) and Boltzmann (B) samples---respectively associated with
    homogenous weights $1$ and $e^\beta$.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist-pk}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
We empirically study the capacity of \ourprog{} to improve the generation of seed sequences for subsequent local optimizations, showing an important application in biologically relevant scenarii.

### 2018-05-04 supplement.tex
Add PK targeting figure; replace deviation by linear fit figure; move pk-free figure to supplemental material

\appendix
{\centering \relsize{+4}Supplementary material\\%
%{\relsize{-1}\BackupTitle}\\
}
\relsize{+1}
\section{Approximate counting and random generation}
In fact, not only is $\#{\sf BIS}$ a reference problem in counting complexity, but it is also a landmark problem with respect to the complexity of approximate counting problems. In this context, it is the representative for a class of $\#{\sf BIS}$-hard problems~\citep{Bulatov2013} that are easier to approximate than $\# {\sf SAT}$, yet are widely believed not to admit any Fully Polynomial-time Randomized Approximation Scheme. Recent results reveal a surprising dichotomy in the behavior of $\#{\sf BIS}$: it admits a Fully Polynomial-Time Approximation Scheme (FPTAS) for graphs of max degree $\le 5$~\citep{Weitz2006}, but is as hard to approximate as the general $\#{\sf BIS}$ problem on graphs of degree $\ge 6$~\citep{Cai2016}. In other words, there is a clear threshold, in term of the max degree, separating (relatively) easy instances from really hard ones.

@@

>>>>

\appendix
{\centering \relsize{+4}\textbf{Supplementary Material}\\ Fixed-Parameter Tractable Sampling\\ for RNA Design with Multiple Target Structures\\%
%{\relsize{-1}\BackupTitle}\\
}
\relsize{+1}

\medskip
\begin{center}
\large Stefan Hammer, Yann Ponty, Wei Wang and Sebastian Will
\end{center}
\section{Approximate counting and random generation}
In fact, not only is $\#{\sf BIS}$ a reference problem in counting complexity, but it is also a landmark problem with respect to the complexity of approximate counting problems. In this context, it is the representative for a class of $\#{\sf BIS}$-hard problems~\citep{Bulatov2013} that are easier to approximate than $\# {\sf SAT}$, yet are widely believed not to admit any Fully Polynomial-time Randomized Approximation Scheme. Recent results reveal a surprising dichotomy in the behavior of $\#{\sf BIS}$: it admits a Fully Polynomial-Time Approximation Scheme (FPTAS) for graphs of max degree $\le 5$~\citep{Weitz2006}, but is as hard to approximate as the general $\#{\sf BIS}$ problem on graphs of degree $\ge 6$~\citep{Cai2016}. In other words, there is a clear threshold, in term of the max degree, separating (relatively) easy instances from really hard ones.

@@

\section{Illustrating multi-dimensional Boltzmann sampling for three pseudoknot-free target structures}

In analogy to Fig.~\ref{fig:energydist-pk} of the main text,
Fig.~\ref{fig:energydist} illustrates our multi-dimensional sampling
strategy for three pseudoknot-free target structures (example instance
from the \texttt{Modena} benchmark). The figure shows Turner energy
distributions of single structures from sampling with different weight
parameters, where weights are trained to target specific energies. For
comparison, the figure shows distributions from uniform and Boltzmann
sampling at respective homogenous weights $1$ and $e^\beta$.

\begin{figure*}[t]
  \begin{center}
    %{\sf \bfseries A}\includegraphicstop[width=0.32\textwidth]{Figs/offset2}\hspace{.5cm}
    %{\sf \bfseries B}
    \includegraphicstop[width=0.8\textwidth]{Figs/energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies using multi-dimensional Boltzmann
    sampling. We show the Turner energy distributions for
    three target structures S1, S2 and S3. By
    targeting the respective free energies $(-40,-40,-20)$ and
    $(-20,-20,-20)$ for the target structures (annotated in red and
    blue, resepectively), we demonstrate the
    effectivity of our adaptive multi-dimensional Boltzmann sampling
    procedure; moreover, for comparison, the distributions for uniform
    (U) and Boltzmann (B) samples---respectively associated with
    homogenous weights $1$ and $e^\beta$.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist}
\end{figure*}
